Appendix

### 2025-03-30 manuscript_rmarkdown/manuscript.Rmd
Reworked robust transformations.

```{=latex}
\begin{equation}
f_{\beta}(x) = \frac{\beta}{2\Gamma\left(1 / \beta \right)} e^{-\left| x \right|^\beta}
\end{equation}
```
Here, $\Gamma$ is the gamma function, and $\beta$ is a strictly positive shape parameter. For $\beta = 1$, the probability density function describes a Laplace distribution. A normal distribution is found for $\beta=2$, and for large $\beta$, the distribution approaches a uniform distribution. We will refrain from introducing scale and location parameters here directly.

@@

\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}

Here, $\alpha \in (0,1)$ is a skewness parameter. $\alpha > 0.5$ creates a distribution with a negative skew, i.e. a left-skewed distribution. A right-skewed distribution is created for $\alpha < 0.5$. $\mu \in \mathcal{R}$ and $\sigma \in (0, \infty)$ are location and scale parameters, respectively. $f_{\alpha}$ thus describes the probability density function of an asymmetric generalised normal distribution, which we will refer to here and parametrise as $\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)$.

We require a quantile function (or an approximation thereof) to draw random values from an asymmetric generalised normal distribution using inverse transform sampling. Gijbels et al. derived the following quantile function $F_{\alpha}^{-1}(p)$:

\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}

The quantile function for the asymmetric generalised normal distribution $F_{\alpha}^{-1}$ thus incorporates the quantile function $F_{\beta}^{-1}$ of the symmetric generalised normal distribution. $F_{\beta}^{-1}$ was derived by Griffin to be [@Griffin2018-bf]:

\begin{equation}
F_{\beta}^{-1}(p) = \sgn\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}

Here, $F_{\Gamma}^{-1}$ is the quantile function of the gamma distribution with shape $1 / \beta$, which can be numerically approximated.

@@

THe alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

@@

\begin{definition}[Central normality test statistic]
Let $\alpha$ be the significance level for the central normality test.
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

@@

We used simulated data to assess invariance to location and scale of the proposed power transformations, weighting for robust transformations, and to develop the empirical central normality test.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well as $\lambda$, $x_0$ and $s$ parameters for location- and scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.
The required algorithms were implemented in the `power.transform` R software package [@Zwanenburg2024-kq] (version 1.0.0).
Of note, the `power.transform` package shifts feature values into the positive domain if negative or zero values are present for Box-Cox power transformations.

@@

To assess whether the proposed power transformations lead to values of $\lambda$ that are invariant to location and scale of the distribution, we simulated three different distributions.
We first randomly drew $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed generalised normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed generalised normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location- and scale-invariant definitions (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

@@

two_sided_function_parameters <- .get_optimised_weighting_function_parameters(
  manuscript_dir = manuscript_dir,
  side = "both"
)

@@

```
## Simulated data

>>>>

\begin{definition}[Standard generalised normal distribution probability density function]
Let $x \in \mathcal{R}$.
Let $\beta > 0$ be a shape parameter.
Let $\Gamma$ be the gamma function.

Then the standard generalised normal distribution probability density function, without scale and location parameters, is:
\begin{equation}
f_{\beta}(x) = \frac{\beta}{2\Gamma\left(1 / \beta \right)} e^{-\left| x \right|^\beta}
\end{equation}
\end{definition}

For $\beta = 1$, the probability density function describes a Laplace distribution.
A normal distribution is found for $\beta=2$, and for large $\beta$,
the distribution approaches a uniform distribution.

@@

\begin{definion}[Asymmetric generalised normal distribution probability density function]
Let $\alpha \in (0,1)$ be a skewness parameter.
Let $\mu \in \mathcal{R}$ be a location parameter.
Let $\sigma > 0$ be a shape parameter.

Then the probability density function of the asymmetric generalised normal distribution is:
\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}
\end{definition}

$\alpha > 0.5$ creates a distribution with a negative skew, i.e. a left-skewed distribution.
A right-skewed distribution is created for $\alpha < 0.5$.
$f_{\alpha}$ thus describes the probability density function of an asymmetric generalised normal distribution,
which we will refer to here and parametrise as $\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)$.

We require a quantile function (or an approximation thereof) to draw random values
from an asymmetric generalised normal distribution using inverse transform sampling.
Gijbels et al. derived the quantile function $F_{\alpha}^{-1}(p)$, which incorporates
the quantile function of the symmetric generalised normal distribution derived by Griffin [@Griffin2018-bf]:

\begin{definition}[Asymmetric generalised normal distribution quantile function]
Let $p \in \left[0, 1\right]$ be a probability.
Let  $F_{\Gamma}^{-1}$ is the quantile function of the gamma distribution with shape $1 / \beta$, which can be numerically approximated. 
Let $F_{\beta}^{-1}$ be the quantile function of the generalised normal distribution:
\begin{equation}
F_{\beta}^{-1}(p) = \sgn\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}

Then the quantile function of the asymmetric generalised normal distribution is:
\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}
\end{definition}

@@

The alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

@@

\begin{definition}[Central normality test statistic]
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

@@

We used simulated data to assess invariance to location and scale of the proposed power transformations,
to develop the empirical central normality test, and to determine weighting for robust transformations.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}),
as well as $\lambda$, $x_0$ and $s$ parameters for location- and scale-invariant power transformations 
(Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm 
for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.
The required algorithms were implemented in the `power.transform` R software package [@Zwanenburg2024-kq] (version 1.0.1).
Of note, the `power.transform` package shifts feature values into the positive domain if negative or zero values are present for Box-Cox power transformations.

@@

To assess whether the proposed power transformations lead to values of $\lambda$ 
that are invariant to location and scale of the distribution, we simulated three different sequences
We first randomly drew $10000$ values from a normal distribution: 
$\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, 
or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed generalised normal distribution 
$\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed generalised normal distribution 
$\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location- and scale-invariant definitions (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

@@

@@

two_sided_function_parameters <- .get_optimised_weighting_function_parameters(
  manuscript_dir = manuscript_dir
)

## Simulated data

### 2025-03-30 manuscript_rmarkdown/manuscript_appendix.Rmd
Reworked robust transformations.

@@

```
```{=latex}
\begin{table}
\begin{center}
\caption{
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:comparison_methods_simulations}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & & 8.51          & 6.33          & 7.31          & 5.55          & 7.42          & 6.39 \\
conventional                          & & 3.72          & 3.78          & 3.83          & 4.74          & 6.47          & 5.47 \\
conventional (z-score norm.)          & & 4.72          & 2.89          & 6.37          & 3.43          & 5.42          & 3.50 \\
conventional (robust scaling)         & & 4.07          & 4.31          & 5.36          & 4.23          & 4.41          & 4.32 \\
Raymaekers-Rousseeuw                  & & 4.50          & 7.03          & 3.97          & 8.12          & 6.25          & 5.57 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 5.26          & 6.36          & 6.22          & 6.42          & 5.26          & 6.92 \\
Raymaekers-Rousseeuw (robust scaling) & & 4.71          & 7.84          & 5.31          & 7.06          & 4.35          & 7.58 \\
invariant                             & & \textbf{3.15} & 3.72          & \textbf{3.27} & 3.14          & \textbf{2.69} & 3.05 \\
robust invariant                      & & 6.37          & \textbf{2.73} & 3.36          & \textbf{2.30} & 2.72          & \textbf{2.20} \\

>>>>

# data[outlier == FALSE, "error" := residuals]
# data[outlier == TRUE, "error" := residuals_central]
# data_baseline <- data[transformation == "conventional"][, mget(c("dataset", "outlier", "experiment_id", "error"))]
# data.table::setnames(data_baseline, "error", "reference_error")
# data <- data[data_baseline, on = c("dataset", "outlier", "experiment_id")]
# data[, "fractional_error" := error / reference_error]

# Add rank based on residual error for central normality.
data[outlier == FALSE, "method_rank" := rank(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "method_rank" := rank(residuals_central), by = c("experiment_id", "outlier", "dataset")]

@@

```{=latex}
\begin{table}
\begin{center}
\caption{
Comparison of average rank between transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.70$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst.
}
\label{tab:comparison_methods_simulations}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         8.45  &         7.52  &         7.29  &         5.99  &         7.39  &         6.39 \\
conventional                          & &         3.64  &         5.57  &         3.78  &         7.04  &         6.45  &         5.79 \\
conventional (z-score norm.)          & &         4.68  &         5.78  &         6.29  &         5.10  &         5.34  &         5.08 \\
conventional (robust scaling)         & &         4.00  &         6.74  &         5.28  &         5.78  &         4.32  &         5.79 \\
Raymaekers-Rousseeuw                  & &         4.41  &         3.66  &         3.92  &         4.40  &         6.23  &         5.50 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.21  &         3.16  &         6.13  &         4.24  &         5.18  &         4.15 \\
Raymaekers-Rousseeuw (robust scaling) & &         4.62  &         3.42  &         5.22  &         4.38  &         4.26  &         4.30 \\
invariant                             & & \textbf{3.05} &         6.95  & \textbf{3.03} &         6.16  & \textbf{2.45} &         6.13 \\
robust invariant                      & &         6.95  & \textbf{2.20} &         4.07  & \textbf{1.90} &         3.38  & \textbf{1.87} \\

### 2025-03-30 manuscript_rmarkdown/manuscript_appendix.tex
Reworked robust transformations.

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-03-27}

@@

>>>>

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-03-30}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\\kappa = 0.70$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst.
}
\label{tab:comparison_methods_simulations}
\begin{tabular}{l | l r r r r r r}

\toprule
& dataset: & \multicolumn{2}{c}{clean} & \multicolumn{2}{c}{dirty} & \multicolumn{2}{c}{shifted} \\
transformation & outliers: & no & yes & no & yes & no & yes \\

none                                  & &         8.45  &         7.52  &         7.29  &         5.99  &         7.39  &         6.39 \\
conventional                          & &         3.64  &         5.57  &         3.78  &         7.04  &         6.45  &         5.79 \\
conventional (z-score norm.)          & &         4.68  &         5.78  &         6.29  &         5.10  &         5.34  &         5.08 \\
conventional (robust scaling)         & &         4.00  &         6.74  &         5.28  &         5.78  &         4.32  &         5.79 \\
Raymaekers-Rousseeuw                  & &         4.41  &         3.66  &         3.92  &         4.40  &         6.23  &         5.50 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.21  &         3.16  &         6.13  &         4.24  &         5.18  &         4.15 \\
Raymaekers-Rousseeuw (robust scaling) & &         4.62  &         3.42  &         5.22  &         4.38  &         4.26  &         4.30 \\
invariant                             & & \textbf{3.05} &         6.95  & \textbf{3.03} &         6.16  & \textbf{2.45} &         6.13 \\
robust invariant                      & &         6.95  & \textbf{2.20} &         4.07  & \textbf{1.90} &         3.38  & \textbf{1.87} \\

\bottomrule
\end{tabular}
\end{center}
\end{table}
